### 2020-07-27 Catuscia - final/06-exampleLactose.tex
da dropbox

% !TEX root = ./main_RS2L_LNCS.tex
\section{Example: \emph{lac} operon}
\label{ex:lactose}
In this section we present the encoding of a RS example
taken from~\cite{CMMBM12}.
%, concerning the lac Operon.
\subsection{The \emph{lac} operon}
An operon is a cluster of genes under the control of a single promoter. 
The \emph{lac} operon  is involved in the metabolism of lactose in \emph{Escherichia coli} cells;
it is composed by three adjacent structural genes (plus some regulatory components):  $lacZ$, $lacY$ and $lacA$ encoding for two enzymes $Z$ and $A$, and a transporter $Y$, involved in the digestion of the $lactose$. The main regulations are:
\begin{itemize}
\item the gene $lacI$ encodes for a repressor protein $I$;
\item the DNA sequence, called \emph{promoter},  is recognised by a RNA polymerase
to iniziate the transcription  of the genes $lacZ$, $lacY$ and $lacA$;
\item a DNA segment, called  the \emph{operator} ($OP$),   obstructs the RNA polymerase functionality when  the repressor protein $I$ is bound to it forming $I\textrm{-}OP$;
\item  a short DNA sequence, called the $CAP\textrm{-}binding\ site$, when it is bound to the complex composed by  the protein $CAP$ and the signal molecule $cAMP$, acts as a promoter for the interaction between the RNA polymerase and the promoter.
\end{itemize}
The functionality of the \emph{lac} operon depends on the integration of two control mechanisms, one mediated by \emph{lactose}, and the other one mediated by \emph{glucose}.

@@

\subsection{The RS formalization}\label{subsec:RSlacoperon}
The reaction system for the \emph{lac} operon is defined as
$A_{lac} = (S, A)$, where the set S represents
the main biochemical components involved in this genetic system, while the reaction set A contains the biochemical
reactions involved in the regulation of the \emph{lac} operon expression. Formally, the \emph{lac} operon reaction system is defined as
follows:
$S$ is the set\\ $\{lac, Z, Y, A, lacI, I, I\textrm{-}OP, cya, cAMP, crp, CAP,cAMP\textrm{-}CAP, lactose,glucose\}$,
and A consists of the following 10 reactions:
{\small
\[
\begin{array}{lcllcl}
a_1 &= &(\{lac\},\{. . .\},\{lac\}), & a_6 &= &(\{cya\},\{. . .\},\{cAMP\}),\\
a_2 &= &(\{lacI\},\{. . .\},\{lacI\}), & a_7 &= &(\{crp\},\{. . .\},\{crp\}),\\
a_3 &= &(\{lacI\},\{. . .\},\{I\}), & a_8 &= &(\{crp\},\{. . .\},\{CAP\}),\\
a_4 &= &(\{I\},\{lactose\},\{I\textrm{-}OP\}), & a_9 &= &(\{cAMP, CAP\},\{glucose\},\{cAMP\textrm{-}CAP\}),\\
a_5 &= &(\{cya\},\{. . .\},\{cya\}),&
a_{10} &=& (\{lac, cAMP\textrm{-}CAP\},\{I\textrm{-}OP\},\{Z, Y, A\}).
\end{array}
\]
}
The default context ($DC$) is composed by those entities that are always present in the system
$DC = \{lac,lacI,I,cya,cAMP,crp,CAP\}$, whereas the $lactose$ and the $glucose$ are given non-deterministically by the context.
\subsection{The RS encoding}
For the sake of readability, the encoding we propose exploits the specific features of the example in hand 
to perform some simplifications:
\begin{itemize}
\item for the entities in the default context, $s \in DC$, as they are persistent, we do not provide the $\overline{P_s}$ processes and the $Cxt_s$ processes;
\item for the reactions requiring the presence of entities $s \in DC$, we do not provide the reaction alternative behaviour for when $s$ is absent;
\item the $Cxt_s$ processes are specified only for those entitiess that are really provided by the context.
\end{itemize}
Moreover, we do not model the \emph{dummy} entity that is specified by  dots ($\dots$) by the RS reactions in Section~\ref{subsec:RSlacoperon}.
Finally, we exclude the \emph{duplication reactions} ($a_1$, $a_2$, $a_5$, $a_7$), and renumber the remaining reactions :
{\small
\[
\begin{array}{cccl}
%a1 &= &(\{lac\},\{. . .\},\{lac\}),\\
%a2 &= &(\{lacI\},\{. . .\},\{lacI\}),\\
\textrm{old } &\textrm{new }&& \textrm{reactions}\\
a_3 &a_1&=&(\{lacI\},\{. . .\},\{I\}),\\
a_4 &a_2&=&(\{I\},\{lactose\},\{I\textrm{-}OP\}),\\
%a5 &= &(\{cya\},\{. . .\},\{cya\}),\\
a_6 & a_3&=&(\{cya\},\{. . .\},\{cAMP\}),\\
%a7 &= &(\{crp\},\{. . .\},\{crp\}),\\
a_8 &a_4&=&(\{crp\},\{. . .\},\{CAP\}),\\
a_9 &a_5&=&(\{cAMP, CAP\},\{glucose\},\{cAMP\textrm{-}CAP\}),\\
a_{10} & a_6&=& (\{lac, cAMP\textrm{-}CAP\},\{I\textrm{-}OP\},\{Z, Y, A\}).
\end{array}
\]
}

%\underline{Processes for the rules}\\
%We provide parametrised code duplication and for expression rules.\\
%\ \\
\paragraph{Duplication reactions.}
As explained previously, their encoding is omitted.
%%$
%\begin{array}{lcccl}
%P_{ai} & \defeq& P_d(ai,s) & = & \startchain{r_i}\chainedlink{s_i}{\noact}\chainedlink{\noact}{s_o}
%\chainedlink{\tilde{s}_i}{\noact}\chainedlink{\noact}{\tilde{s}_o}\chainend{r_{i+1}}.P_d(ai,s)
%\end{array}
%$,
%with $r_1 = \silent$.
%There are four duplication rules:\\ $P_d(a1,lac)$, $P_d(a2,lacI)$,$P_d(a5,cya)$,$P_d(a7,crp)$.\\
\paragraph{Expression reactions.}
First we define the parametric process
\[
\begin{array}{lcl}
% P_i(s1,s2) & \defeq & \startchain{r_i}\chainedlink{s1_i}{\noact}\chainedlink{\noact}{s1_o}
%\chainedlink{\tilde{s2}_i}{\noact}\chainedlink{\noact}{\tilde{s2}_o}\chainend{r_{i+1}}.P_i(s1,s2)\\
 P_i(s1,s2) & \defeq & \startchain{r_i}\chainedlink{s1_i}{\noact}\chainedlink{\noact}{s1_o}
 \chainedlink{r_{i+1}}{\noact}\chainedlink{\noact}{p_{i}}\chainedlink{\widetilde{s2}_i}{\noact}
\chainedlink{\noact}{\widetilde{s2}_o}\chainend{p_{i+1}}.P_i(s1,s2)\\
\end{array}
\]
Then, we let $P_{a1}\defeq P_1(lacI,I)$,  $P_{a3} \defeq P_3(cya,cAMP)$, and $P_{a4} \defeq P_4(crp,CAP)$.
%\[
%\begin{array}{lcl}
%P_{a1} & =& \startchain{\silent}\chainedlink{lac_i}{\noact}\chainedlink{\noact}{lac_o}\chainedlink{\tilde{lac}_i}{\noact}\chainedlink{\noact}{\tilde{lac}_o}\chainend{r_2}.P_{a1}\\
%P_{a2} & =& \startchain{r_2}\chainedlink{lacI_i}{\noact}\chainedlink{\noact}{lacI_o}\chainedlink{\tilde{lacI}_i}{\noact}\chainedlink{\noact}{\tilde{lacI}_o}\chainend{r_3}.P_{a2}\\
%P_{a3} & =& \startchain{r_3}\chainedlink{I_i}{\noact}\chainedlink{\noact}{I_o}\chainedlink{\tilde{I}_i}{\noact}\chainedlink{\noact}{\tilde{I}_o}\chainend{r_4}.P_{a3}\\
\paragraph{Regulation reactions.}

@@

@@

\chainedlink{cAMP_i}{\noact}\chainedlink{\noact}{cAMP_o}\chainedlink{\overline{I\textrm{-}OP}_i}{\noact}

@@

\chainedlink{\widetilde{z}_i}{\noact}\chainedlink{\noact}{\widetilde{z}_o}
\chainedlink{\widetilde{y}_i}{\noact}\chainedlink{\noact}{\widetilde{y}_o}
\chainedlink{\widetilde{A}_i}{\noact}\chainedlink{\noact}{\widetilde{A}_o}

@@

&&  \startchain{r_{6}}\chainedlink{\overline{lac}_i}{\noact}\chainedlink{\noact}{\overline{lac}_o} \chainedlink{cxt_1}{\noact}\chainedlink{\noact}{p_6}\chainend{\silent}.P_{a6} \ + \ \chainedlink{\overline{cAMP}_i}{\noact}\chainedlink{\noact}{\overline{cAMP}_o} \chainedlink{cxt_1}{\noact}\chainedlink{\noact}{p_6}\chainend{\silent}.P_{a6}

@@

We exploit the specificity of the example in hand to optimising the code, and we specify exactly

@@

@@

%P_{lac}&\defeq&\startchain{lac_i}\chainedlink{lac_o}{\noact}\chainedlink{\noact}{lac_i}\chainedlink{lac_o}{\noact}\chainedlink{\noact}{\tilde{lac}_i}\chainend{\tilde{lac}_o}.P_{lac} +
 %\startchain{lac_i}\chainedlink{lac_o}{\noact}\chainedlink{\noact}{\tilde{lac}_i}\chainend{\tilde{lac}_o}.P_{lac}\\
% P_{lac}&\defeq&\sum_{h=1}^2(\startchain{lac_i}\chainedlink{lac_o}{\noact}\chainend{\noact})^h\ \link{\tilde{lac}_i}{\tilde{lac}_o}.P_{lac}
%P_{cya}&\defeq&\startchain{cya_i}\chainedlink{cya_o}{\noact}\chainedlink{\noact}{cya_i}\chainedlink{cya_o}{\noact}
%\chainedlink{\noact}{\tilde{cya}_i}\chainend{\tilde{cya}_o}.P_{cya}\\

@@

%&
%P_{I}&\defeq&\sum_{h=0}^1(\startchain{I_i}\chainedlink{I_o}{\noact}\chainend{\noact})^h\ \link{\tilde{I}_i}{\tilde{I}_o}.P_{I} \\
 %{\noact}
%\chainedlink{\noact}{\cap{lac}_i}\chainend{\cap{lac}_o}.P_{lac}\\
%P_{crp}&\defeq&\startchain{crp_i}\chainedlink{crp_o}{\noact}\chainedlink{\noact}{crp_i}\chainedlink{crp_o}{\noact}\chainedlink{\noact}{\tilde{crp}_i}\chainend{\tilde{crp}_o}.P_{crp}

@@

%&
%P_{cAMP}&\defeq&\sum_{h=0}^1(\startchain{cAMP_i}\chainedlink{cAMP_o}{\noact}\chainend{\noact})^h\ \link{\widetilde{cAMP}_i}{\widetilde{cAMP}_o}.P_{cAMP}

@@

%&
%P_{CAP}&\defeq&\sum_{h=0}^1(\startchain{CAP_i}\chainedlink{CAP_o}{\noact}\chainend{\noact})^h\ \link{\widetilde{CAP}_i}{\widetilde{CAP}_o}.P_{CAP}\\
%{lacI_i}\chainedlink{lacI_o}{\noact}\chainedlink{\noact}{\tilde{lacI}_i}\chainend{\tilde{lacI}_o}.P_{lacI}\\
% \startchain{\tilde{I}_i}\chainend{\tilde{I}_o}.P_{I}\\
%P_{I}&\defeq&\startchain{I_i}\chainedlink{I_o}{\noact}\chainedlink{\noact}{\tilde{I}_i}\chainend{\tilde{I}_o}.P_{I} +
% \startchain{\tilde{I}_i}\chainend{\tilde{I}_o}.P_{I}\\
%P_{cya}&\defeq&\startchain{cya_i}\chainedlink{cya_o}{\noact}\chainedlink{\noact}{cya_i}\chainedlink{cya_o}{\noact}\chainedlink{\noact}{\tilde{cya}_i}\chainend{\tilde{cya}_o}.P_{cya}\\
%P_{lacI}&\defeq&\sum_{h=1}^2(\startchain{lacI_i}\chainedlink{lacI_o}{\noact}\chainend{\noact})^h\ \link{\tilde{lacI}_i}{\tilde{lacI}_o}.P_{lacI}\\

@@

%P_{cAMP}&\defeq&\startchain{cAMP_i}\chainedlink{cAMP_o}{\noact}\chainedlink{\noact}{\tilde{cAMP}_i}\chainend{\tilde{cAMP}_o}.P_{cAMP} + \startchain{\tilde{cAMP}_i}\chainend{\tilde{cAMP}_o}.P_{cAMP}\\
%P_{crp}&\defeq&\startchain{crp_i}\chainedlink{crp_o}{\noact}\chainedlink{\noact}{crp_i}\chainedlink{crp_o}{\noact}\chainedlink{\noact}{\tilde{crp}_i}\chainend{\tilde{crp}_o}.P_{crp}\\

@@

$P_e(s) \defeq  \startchain{s_i}\chainedlink{s_o}{\noact}\chainedlink{\noact}{\widetilde{s}_i}\chainend{\widetilde{s}_o}.P_e(s) +  \startchain{\widetilde{s_i}}\chainend{\widetilde{s}_o} .P_e(s) $.\\

@@

The entity $I\textrm{-}OP$ can be either produced (by $a_2$) or tested for absence (by $a_6$).
Correspondingly, the process  $P_{I\textrm{-}OP}$ is defined as follows:

\[\begin{array}{lcl}
P_{I\textrm{-}OP}&\defeq & 
\sum_{h=0}^1 (\startchain{I\textrm{-}OP_i}\chainedlink{I\textrm{-}OP_o}{\noact}\chainend{\noact})^h  \, \startchain{\widetilde{I\textrm{-}OP}_i}\chainend{\widetilde{I\textrm{-}OP}_o}.P_{I\textrm{-}OP}
%\\
%&&+\\
%&&\sum_{h=0}^1(\startchain{I-OP_i}\chainedlink{I-OP_o}{\noact}\chainend{\noact})^h\ \startchain{\widetilde{I-OP}_i}\chainedlink{\widetilde{I-OP}_o}{\noact}\chainedlink{\noact}{\underline{I-OP}_i}\chainend{\underline{I-OP}_o}.P_{I-OP}\\
%&&
\  +\ %\\
%&&
\link{I\textrm{-}OP_i}{I\textrm{-}OP_o}.\overline{P_{I\textrm{-}OP}}\\
\overline{P_{I\textrm{-}OP}} & \defeq & 
% P_{I\textrm{-}OP}(\overline{I\textrm{-}OP})\\
\sum_{h=0}^1  (\startchain{\overline{I\textrm{-}OP}_i}\chainedlink{\overline{I\textrm{-}OP}_o}{\noact}\chainend{\noact})^h \,
\startchain{\widetilde{I\textrm{-}OP}_i}\chainend{\widetilde{I\textrm{-}OP}_o}.P_{I\textrm{-}OP}%\\
%&&
\ +\ %\\
%&&
\link{\overline{I\textrm{-}OP}_i}{\overline{I\textrm{-}OP}_o}.\overline{P_{I\textrm{-}OP}}\\

\end{array}
\]
\noindent
The  process $P_{cAMP\textrm{-}CAP}$ is similar to $P_{I\textrm{-}OP}$, as it is produced by $a_5$ and tested for presence by $a_6$. Its code is  in Table~\ref{tab:cAMP-CAP}, in the Appendix.
The $lactose$ is provided by the context and tested for absence by $a_2$.

\[
\begin{array}{lcllcl}
P_{lactose} &\defeq &%== \overline{P_{glucose}}&\defeq & P_{glu-lact}(\overline{glucose})\\
\sum_{h=0}^1 (\startchain{lactose_i}\chainedlink{lactose_o}{\noact}\chainend{\noact})^h\ \link{\widehat{lactose_i}}{\widehat{lactose_o}}.P_{lactose}
\\
&& + \\
&& \sum_{h=0}^1 (\startchain{lactose_i}\chainedlink{lactose_o}{\noact}\chainend{\noact})^h\ \link{\underline{lactose_i}}{\underline{lactose_o}}.\overline{P_{lactose}}\\
%P_{glucose} &\defeq &\sum_{h=0}^1 \underbrace{\link{glucose_i}{glucose_o}}_{h\ times}
%
%\[
%\begin{array}{lcl}

\overline{P_{lactose}} &\defeq &\sum_{h=0}^1 (\startchain{\overline{lactose}_i}\chainedlink{\overline{lactose}_o}{\noact}\chainend{\noact})^h\ \link{\widehat{lactose_i}}{\widehat{lactose_o}}.P_{lactose}\\
&&+\\
&& \sum_{h=0}^1 (\startchain{\overline{lactose}_i}\chainedlink{\overline{lactose}_o}{\noact}\chainend{\noact})^h\ \link{\underline{lactose_i}}{\underline{lactose_o}}.\overline{P_{lactose}}\\
\end{array}
\]
\noindent
The  process $P_{glucose}$ is similar to $P_{lactose}$ and tested for absence by $a_5$. Its code is  in Table~\ref{tab:glucose}, in the Appendix.
The entity $z$ can only be produced by rule $a_6$, while it is never provided by the context. Moreover, there is no rule for testing its presence or absence.
\[
\begin{array}{lcl@{\qquad\qquad}lcl}
P_z&\defeq &\startchain{\widetilde{z_i}}\chainedlink{\widetilde{z_o}}{\noact}\chainedlink{\noact}{\underline{z_i}}\chainend{\underline{z_o}}.P_z\ + \
\link{\underline{z_i}}{\underline{z_o}}.\overline{P_z}
&
\overline{P_z} &\defeq &\startchain{\widetilde{z_i}}\chainedlink{\widetilde{z_o}}{\noact}\chainedlink{\noact}{\underline{z_i}}\chainend{\underline{z_o}}.\overline{P_z}\ + \
\link{\underline{z_i}}{\underline{z_o}}.\overline{P_z}\\\end{array}
\]
\noindent
The entities $y$ and $A$ are treated in the same way as $z$. Their processes are in Table~\ref{tab:yA} in the Appendix.

>>>>

% !TEX root = ./main_RS2L_LNCS.tex
\section{Example: \emph{lac} operon}
\label{ex:lactose}
In this section we present the encoding of a RS example
taken from~\cite{CMMBM12}.
%, concerning the lac Operon.
\subsection{The \emph{lac} operon}
An operon is a cluster of genes under the control of a single promoter. 
The \emph{lac} operon  is involved in the metabolism of lactose in \emph{Escherichia coli} cells;
it is composed by three adjacent structural genes (plus some regulatory components):  $lacZ$, $lacY$ and $lacA$ encoding for two enzymes $Z$ and $A$, and a transporter $Y$, involved in the digestion of the $lactose$. The main regulations are:
\begin{itemize}
\item 
%\begin{color}{red}
the gene $lacI$ encodes a repressor protein $I$;
%\end{color}
\item the DNA sequence, called \emph{promoter},  is recognised by a RNA polymerase
to iniziate the transcription  of the genes $lacZ$, $lacY$ and $lacA$;
\item a DNA segment, called  the \emph{operator} ($OP$),   obstructs the RNA polymerase functionality when  the repressor protein $I$ is bound to it forming $I\textrm{-}OP$;
\item  a short DNA sequence, called the $CAP\textrm{-}binding\ site$, when it is bound to the complex composed by  the protein $CAP$ and the signal molecule $cAMP$, acts as a promoter for the interaction between the RNA polymerase and the promoter.
\end{itemize}
The functionality of the \emph{lac} operon depends on the integration of two control mechanisms, one mediated by \emph{lactose}, and the other one mediated by \emph{glucose}.

@@

\subsection{The RS formalization}\label{subsec:RSlacoperon}
The reaction system for the \emph{lac} operon is defined as
$A_{lac} = (S, A)$, where the set S represents
the main biochemical components involved in this genetic system, while the reaction set A contains the biochemical
reactions involved in the regulation of the \emph{lac} operon expression. Formally, the \emph{lac} operon reaction system is defined as
follows:
$S$ is the set
$$\{lac, Z, Y, A, lacI, I, I\textrm{-}OP, cya, cAMP, crp, CAP,cAMP\textrm{-}CAP, lactose,glucose\},$$
and A consists of the following 10 reactions:
\begin{array}{lcllcl}
a_1 &= &(\{lac\},\{. . .\},\{lac\}), & a_6 &= &(\{cya\},\{. . .\},\{cAMP\}),\\
a_2 &= &(\{lacI\},\{. . .\},\{lacI\}), & a_7 &= &(\{crp\},\{. . .\},\{crp\}),\\
a_3 &= &(\{lacI\},\{. . .\},\{I\}), & a_8 &= &(\{crp\},\{. . .\},\{CAP\}),\\
a_4 &= &(\{I\},\{lactose\},\{I\textrm{-}OP\}), & a_9 &= &(\{cAMP, CAP\},\{glucose\},\{cAMP\textrm{-}CAP\}),\\
a_5 &= &(\{cya\},\{. . .\},\{cya\}),&
a_{10} &=& (\{lac, cAMP\textrm{-}CAP\},\{I\textrm{-}OP\},\{Z, Y, A\}).
The default context ($DC$) is composed by those entities that are always present in the system
$DC = \{lac,lacI,I,cya,cAMP,crp,CAP\}$, whereas the $lactose$ and the $glucose$ are given non-deterministically by the context.
\subsection{The RS encoding}
For the sake of readability, the encoding we propose exploits the specific features of the example in hand 
to perform some simplifications:
\begin{itemize}
\item for the entities in the default context, $s \in DC$, as they are persistent, we do not provide the $\overline{P_s}$ processes and the $Cxt_s$ processes;
\item for the reactions requiring the presence of entities $s \in DC$, we do not provide the reaction alternative behaviour when $s$ is absent;
\item the $Cxt_s$ processes are specified only for those entitiess that are really provided by the context.
\end{itemize}
Moreover, we do not model the \emph{dummy} entity that is specified by  dots ($\dots$) by the RS reactions in Section~\ref{subsec:RSlacoperon}.
Finally, we exclude the \emph{duplication reactions} ($a_1$, $a_2$, $a_5$, $a_7$), and renumber the remaining reactions :
\begin{array}{cccl}
%a1 &= &(\{lac\},\{. . .\},\{lac\}),\\
%a2 &= &(\{lacI\},\{. . .\},\{lacI\}),\\
\textrm{old } &\textrm{new }&& \textrm{reactions}\\
a_3 &a_1&=&(\{lacI\},\{. . .\},\{I\}),\\
a_4 &a_2&=&(\{I\},\{lactose\},\{I\textrm{-}OP\}),\\
%a5 &= &(\{cya\},\{. . .\},\{cya\}),\\
a_6 & a_3&=&(\{cya\},\{. . .\},\{cAMP\}),\\
%a7 &= &(\{crp\},\{. . .\},\{crp\}),\\
a_8 &a_4&=&(\{crp\},\{. . .\},\{CAP\}),\\
a_9 &a_5&=&(\{cAMP, CAP\},\{glucose\},\{cAMP\textrm{-}CAP\}),\\
a_{10} & a_6&=& (\{lac, cAMP\textrm{-}CAP\},\{I\textrm{-}OP\},\{Z, Y, A\}).
%\paragraph{Duplication reactions.}
%As explained previously, their encoding is omitted.
%%$
%\begin{array}{lcccl}
%P_{ai} & \defeq& P_d(ai,s) & = & \startchain{r_i}\chainedlink{s_i}{\noact}\chainedlink{\noact}{s_o}
%\chainedlink{\tilde{s}_i}{\noact}\chainedlink{\noact}{\tilde{s}_o}\chainend{r_{i+1}}.P_d(ai,s)
%$,
%with $r_1 = \silent$.
%There are four duplication rules:\\ $P_d(a1,lac)$, $P_d(a2,lacI)$,$P_d(a5,cya)$,$P_d(a7,crp)$.\\
\paragraph{Expression reactions.}
First we define the parametric process
% P_i(s1,s2) & \defeq & \startchain{r_i}\chainedlink{s1_i}{\noact}\chainedlink{\noact}{s1_o}
%\chainedlink{\tilde{s2}_i}{\noact}\chainedlink{\noact}{\tilde{s2}_o}\chainend{r_{i+1}}.P_i(s1,s2)\\
 P_i(s1,s2) & \defeq & \startchain{r_i}\chainedlink{s1_i}{\noact}\chainedlink{\noact}{s1_o}
 \chainedlink{r_{i+1}}{\noact}\chainedlink{\noact}{p_{i}}\chainedlink{\tilde{s2}_i}{\noact}
\chainedlink{\noact}{\tilde{s2}_o}\chainend{p_{i+1}}.P_i(s1,s2)\\
Then, we let $P_{a1}\defeq P_1(lacI,I)$,  $P_{a3} \defeq P_3(cya,cAMP)$, and $P_{a4} \defeq P_4(crp,CAP)$.
%\[
%P_{a1} & =& \startchain{\silent}\chainedlink{lac_i}{\noact}\chainedlink{\noact}{lac_o}\chainedlink{\tilde{lac}_i}{\noact}\chainedlink{\noact}{\tilde{lac}_o}\chainend{r_2}.P_{a1}\\
%P_{a2} & =& \startchain{r_2}\chainedlink{lacI_i}{\noact}\chainedlink{\noact}{lacI_o}\chainedlink{\tilde{lacI}_i}{\noact}\chainedlink{\noact}{\tilde{lacI}_o}\chainend{r_3}.P_{a2}\\
%P_{a3} & =& \startchain{r_3}\chainedlink{I_i}{\noact}\chainedlink{\noact}{I_o}\chainedlink{\tilde{I}_i}{\noact}\chainedlink{\noact}{\tilde{I}_o}\chainend{r_4}.P_{a3}\\
\paragraph{Regulation reactions.}

@@

 \\

@@

\chainedlink{cAMP\textrm{-}CAP_i}{\noact}\chainedlink{\noact}{cAMP\textrm{-}CAP_o}\chainedlink{\overline{I\textrm{-}OP}_i}{\noact}

@@

\chainedlink{\tilde{z}_i}{\noact}\chainedlink{\noact}{\tilde{z}_o}
\chainedlink{\tilde{y}_i}{\noact}\chainedlink{\noact}{\tilde{y}_o}
\chainedlink{\tilde{A}_i}{\noact}\chainedlink{\noact}{\tilde{A}_o}

@@

&&  \startchain{r_{6}}\chainedlink{\overline{lac}_i}{\noact}\chainedlink{\noact}{\overline{lac}_o} \chainedlink{cxt_1}{\noact}\chainedlink{\noact}{p_6}\chainend{\silent}.P_{a6} \ + \ \chainedlink{\overline{cAMP\textrm{-}CAP}_i}{\noact}\chainedlink{\noact}{\overline{cAMP\textrm{-}CAP}_o} \chainedlink{cxt_1}{\noact}\chainedlink{\noact}{p_6}\chainend{\silent}.P_{a6}

@@

We exploit the specificity of the example in hand to optimise the code, and we specify exactly

@@

%{\color{red} For the always present entities we define the parametric process $P(s)\defeq \link{s_i}{s_o}.P_{s}$, and we let:
%\[
%\begin{array}{lcl@{\qquad}lcl@{\qquad}lcl@{\qquad}lcl}
%P_{cya}&\defeq&P(cya) &P_{crp}&\defeq&P(crp)&P_{lacI}&\defeq&P(lacI) &
% P_{lac}&\defeq&P(lac)
%\end{array}\]

@@

@@

@@

@@

@@

@@

$P_e(s) \defeq  \startchain{s_i}\chainedlink{s_o}{\noact}\chainedlink{\noact}{\tilde{s}_i}\chainend{\tilde{s}_o}.P_e(s) +  \startchain{\tilde{s_i}}\chainend{\tilde{s}_o} .P_e(s) $.\\

@@

The entity $I\textrm{-}OP$ can be either produced (by $a_2$) or tested for absence (by $a_6$).
Correspondingly, the process  $P_{I\textrm{-}OP}$ is defined as follows:
\[\begin{array}{lcl}
P_{I\textrm{-}OP}&\defeq & 
\sum_{h=0}^1 (\startchain{I\textrm{-}OP_i}\chainedlink{I\textrm{-}OP_o}{\noact}\chainend{\noact})^h  \, \startchain{\widetilde{I\textrm{-}OP}_i}\chainend{\widetilde{I\textrm{-}OP}_o}.P_{I\textrm{-}OP}
%\\
%&&+\\
%&&\sum_{h=0}^1(\startchain{I-OP_i}\chainedlink{I-OP_o}{\noact}\chainend{\noact})^h\ \startchain{\widetilde{I-OP}_i}\chainedlink{\widetilde{I-OP}_o}{\noact}\chainedlink{\noact}{\underline{I-OP}_i}\chainend{\underline{I-OP}_o}.P_{I-OP}\\
%&&
\  +\ %\\
%&&
\link{I\textrm{-}OP_i}{I\textrm{-}OP_o}.\overline{P_{I\textrm{-}OP}}\\
\overline{P_{I\textrm{-}OP}} & \defeq & 
% P_{I\textrm{-}OP}(\overline{I\textrm{-}OP})\\
\sum_{h=0}^1  (\startchain{\overline{I\textrm{-}OP}_i}\chainedlink{\overline{I\textrm{-}OP}_o}{\noact}\chainend{\noact})^h \,
\startchain{\widetilde{I\textrm{-}OP}_i}\chainend{\widetilde{I\textrm{-}OP}_o}.P_{I\textrm{-}OP}%\\
%&&
\ +\ %\\
%&&
\link{\overline{I\textrm{-}OP}_i}{\overline{I\textrm{-}OP}_o}.\overline{P_{I\textrm{-}OP}}\\

\end{array}
\]
\noindent
The  process $P_{cAMP\textrm{-}CAP}$ is similar to $P_{I\textrm{-}OP}$, as it is produced by $a_5$ and tested for presence by $a_6$. 
Its code is omitted.
%Its code is  in Table~\ref{tab:cAMP-CAP}, in the Appendix.

The $lactose$ is provided by the context and tested for absence by $a_2$.

\[
\begin{array}{lcllcl}
P_{lactose} &\defeq &%== \overline{P_{glucose}}&\defeq & P_{glu-lact}(\overline{glucose})\\
\sum_{h=0}^1 (\startchain{lactose_i}\chainedlink{lactose_o}{\noact}\chainend{\noact})^h\ \link{\widehat{lactose_i}}{\widehat{lactose_o}}.P_{lactose}
\\
&& + \\
&& \sum_{h=0}^1 (\startchain{lactose_i}\chainedlink{lactose_o}{\noact}\chainend{\noact})^h\ \link{\underline{lactose_i}}{\underline{lactose_o}}.\overline{P_{lactose}}\\
\\
\overline{P_{lactose}} &\defeq &\sum_{h=0}^1 (\startchain{\overline{lactose}_i}\chainedlink{\overline{lactose}_o}{\noact}\chainend{\noact})^h\ \link{\widehat{lactose_i}}{\widehat{lactose_o}}.P_{lactose}\\
&&+\\
&& \sum_{h=0}^1 (\startchain{\overline{lactose}_i}\chainedlink{\overline{lactose}_o}{\noact}\chainend{\noact})^h\ \link{\underline{lactose_i}}{\underline{lactose_o}}.\overline{P_{lactose}}\\
\end{array}
\]
\noindent
The  process $P_{glucose}$ is similar to $P_{lactose}$ and tested for absence by $a_5$. 
Its code is omitted.
%Its code is  in Table~\ref{tab:glucose}, in the Appendix.

The entity $z$ can only be produced by rule $a_6$, while it is never provided by the context. Moreover, there is no rule for testing its presence or absence.
\[
\begin{array}{lcl@{\qquad\qquad}lcl}
P_z&\defeq &\startchain{\tilde{z_i}}\chainedlink{\tilde{z_o}}{\noact}\chainedlink{\noact}{\underline{z_i}}\chainend{\underline{z_o}}.P_z\ + \
\link{\underline{z_i}}{\underline{z_o}}.\overline{P_z}
&
\overline{P_z} &\defeq &\startchain{\tilde{z_i}}\chainedlink{\tilde{z_o}}{\noact}\chainedlink{\noact}{\underline{z_i}}\chainend{\underline{z_o}}.\overline{P_z}\ + \
\link{\underline{z_i}}{\underline{z_o}}.\overline{P_z}\\\end{array}
\]
\noindent
The entities $y$ and $A$ are treated likewise $z$. Their processes are omitted.
%in Table~\ref{tab:yA} in the Appendix.

\paragraph{Context.}
%Now, we define the behaviour of the context.
The entities in $DC$ are assumed always present by default, so no context process is needed for them.
The entities $z$, $y$, and $A$ are assumed never provided by the context:
% Their processes are
\[
\begin{array}{rcl@{\qquad}rcl}
Cxt_{z}&\defeq & \startchain{cxt_1}\chainedlink{\underline{z}_i}{\noact}\chainedlink{\noact}{\underline{z}_o}\chainend{cxt_2}.Cxt_{z}
&
Cxt_{y}&\defeq & \startchain{cxt_2}\chainedlink{\underline{y}_i}{\noact}\chainedlink{\noact}{\underline{y}_o}\chainend{cxt_3}.Cxt_{y}
\\
Cxt_{A}&\defeq & \startchain{cxt_3}\chainedlink{\underline{A}_i}{\noact}\chainedlink{\noact}{\underline{A}_o}\chainend{cxt_4}.Cxt_{A} 
\end{array}
\]
 Also, for the sake of presentation, we assume that the $glucose$ is never provided and $lactose$ is always provided by the context:
\[
\begin{array}{lcl}
Cxt_{lactose}&\defeq& \startchain{cxt_4}\chainedlink{\widehat{lactose}_i}{\noact}\chainedlink{\noact}{\widehat{lactose}_o}\chainend{cxt_5}.Cxt_{lactose}\\
Cxt_{glucose}&\defeq & \startchain{cxt_5}\chainedlink{\underline{glucose}_i}{\noact}\chainedlink{\noact}{\underline{glucose}_o}\chainend{p_1}.Cxt_{glucose}
\end{array}
\]

\noindent
%For the sake of presentation, we assume that the $lactose$ is always provided by the context, in contrast,
%$glucose$ is never provided.
%\[
%\begin{array}{lcllcl}
%Cxt_{lactose}&\defeq& \startchain{cxt_4}\chainedlink{\widehat{lactose}_i}{\noact}\chainedlink{\noact}{\widehat{lactose}_o}\chainend{cxt_5}.Cxt_{lactose}
%\\
%Cxt_{glucose}&\defeq & \startchain{cxt_5}\chainedlink{\underline{glucose}_i}{\noact}\chainedlink{\noact}{\underline{glucose}_o}\chainend{p_1}.Cxt_{glucose}
%\end{array}
%\]
\noindent
In the following we let $CXT \defeq Cxt_z | Cxt_y |Cxt_A | Cxt_{lactose} | Cxt_{glucose}$ be the processes for context. 
%
The whole system
%, encoding an extended interactive process, with the previous context behaviour
is as follows: 
\[ lacOp \defeq \restrict{names} (\Pi_{i=1}^6 P_{ai} | \Pi_{s \in DC} P_s |   \Pi_{s \in S \backslash DC} \overline{P}_s| CXT)
\]
